Live Functional Programming with Typed Holes

### 2018-02-07 hazelnut-dynamics.tex
tightening up the abstract

%% Abstract
%% Note: \begin{abstract}...\end{abstract} environment must come
%% before \maketitle command
\begin{abstract}
\emph{Live programming environments} are designed to provide programmers (and, in some cases, audiences) with rapid and continuous feedback about a program's dynamic behavior throughout the editing process. The problem is that programming languages typically assign dynamic meaning only to programs that are syntactically well-formed and free of static type errors (collectively, \emph{complete programs}), whereas programming environments often encounter edit states that do not correspond to complete programs. Recent work by \citet{popl-paper} has made progress on this problem of working with incomplete programs by developing a type system for lambda terms with \emph{holes} in both expression and type position, with empty holes standing for missing expressions or types, and non-empty holes operating as ``membranes'' around type inconsistencies. This paper contributes a complementary dynamic semantics, i.e. one that assigns dynamic meaning to every incomplete program accepted by the type system in the prior work. Of course, it would not be entirely satisfying to assign trivial dynamic meaning to incomplete programs, e.g. by simply raising an exception as soon as evaluation first encounters a hole (as in some existing systems). Instead, evaluation proceeds ``around'' expression holes, performing as much computation as is possible while also tracking the dynamic environment around each expression hole as it flows through the program. These hole environments can be reported directly to the programmer as they work to fill the hole, and they also support an ``edit-and-resume'' feature for edits that amount to hole filling, rather than requiring that evaluation restart after each such edit. Formally, our semantics incorporates features from both gradual type theory (to support evaluation of terms with type holes) and contextual modal type theory (CMTT, to support environment tracking), combined with technical novelties necessary to maintain type safety in the presence of expression holes (i.e. free metavariables, from the perspective of CMTT) and prove the commutativity and confluence properties that guarantee that the ``edit-and-resume'' feature is sound. The primary focus of this paper is on these theoretical properties, which we have mechanized using the Agda proof assistant. We additionally describe a simple reference implementation, called Hazelnut Live, that exposes a language of structured edit actions based on those defined in the prior work by \citet{popl-paper} that insert holes as necessary to guarantee that every edit state has some (possibly incomplete) type. This property, together with the type safety property we prove, results in a live functional programming environment where feedback never lags the edit state, i.e. \emph{every edit state is guaranteed to have non-trivial static and dynamic meaning}.

>>>>

%% Abstract
%% Note: \begin{abstract}...\end{abstract} environment must come
%% before \maketitle command
\begin{abstract}
\emph{Live programming environments} aim to provide programmers (and, in many cases, audiences) with continuous feedback about a program's dynamic behavior throughout the editing process. The problem that motivates this paper is that programming languages typically assign dynamic meaning only to programs that are syntactically well-formed and free of static type errors (collectively, \emph{complete programs}), whereas program editors frequently encounter edit states that are incomplete. Recent work by \citet{popl-paper} has made progress on the problem of working directly with incomplete programs by developing a type system for lambda terms with \emph{holes} in both expression and type position, with empty holes standing for missing expressions or types, and non-empty holes operating as ``membranes'' around type inconsistencies. This paper develops a corresponding dynamic semantics. Rather than aborting with an exception when evaluation encounters a hole (as in  several existing systems), evaluation proceeds ``around'' expression holes, performing as much computation as is possible and tracking the dynamic environment around each hole as it flows through the program. These accumulated hole environments can be reported directly to the programmer by the editor as they work to fill holes. Hole environments also enable an ``edit-and-resume'' feature that avoids the need to restart evaluation after edits that amount to hole filling. Formally, the semantics combines features from gradual type theory (to handle incomplete types) and contextual modal type theory (CMTT, which provides a logical foundation for hole environment tracking) with novel structure necessary to prove key metatheoretic properties, including 1) a type safety property that accounts for expression holes (i.e. free metavariables, from the perspective of CMTT), and 2) commutativity and confluence properties that guarantee that the ``edit-and-resume'' feature is sound. The primary focus of this paper is on these formal properties, which we have mechanized using the Agda proof assistant. We additionally describe a simple reference implementation, called Hazelnut Live, that exposes a language of structured edit actions that insert holes as necessary to guarantee that every edit state has some (possibly incomplete) type, based on the edit action calculus in the prior work by \citet{popl-paper}. Taken together, the result is a live functional programming environment where feedback is truly continuous, i.e. it is available for every possible edit state. % i.e. \emph{every edit state is guaranteed to have non-trivial static and dynamic meaning}.
